Nice guys or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations

### 2019-07-29 article_english.Rmd
Translation

2. There will be an effect of social closeness on cooperation, regardless of promises. We expect that there is more probability of paying the _ friend_ than the _ stranger_ and more probability of paying the stranger than the _computer_. According to the _instrumental_ motivation, the cooperation will be greater for partners with whom the subject anticipates greater cooperation in the future (friend> stranger> computer).

@@

@@

>>>>

2. There will be an effect of social closeness on cooperation, regardless of promises. We expect that there is more probability of paying the _friend_ than the _stranger_ and more probability of paying the stranger than the _computer_. According to the _instrumental_ motivation, the cooperation will be greater for partners with whom the subject anticipates greater cooperation in the future (friend > stranger > computer).

@@

## Cumulative Bayesian Modeling

@@

The objective of this study was to evaluate the effect of social closeness on keeping and breaking promises, as well as on cooperation in a trust game. Our results give evidence that zero social closeness increases the probability of breaking the promise and it decreases monotonically as social closeness increases. Likewise, as social closeness increases, so does cooperation; _high_ social closeness has an effect on decisions that surpasses even those that all other experimental conditions. The most interesting finding was that when subjects express high commitment to cooperate (through choosing that they would _always_ pay back) probability that they comply is also quite high, however, there is a proportion of transgressions that according to our inferences are between 4 - 38% with 95% certainty.

To our knowledge, this is the first experimental study that incorporates social closeness as a predictor of the decision to break promises with socially relevant partners. In previous studies, the participants remain anonymous during the course of the tasks [@Baumgartner2009], their measurements to evaluate transgressions are self-reported [@DePaulo1998; @DePaulo2004], they do not directly quantify the breaking of promises [@Vanberg2008; @Charness2006], or as we will discuss later, they use heterogeneous measures of social closeness [@Glaeser2000].

Our experiment is probably the first to include a partner without social closeness (_computer_), our data give evidence that the mere fact of considering that you play with a human diminishes the probability of breaking the promise. It is approximately 7.5 times more likely to break the promise to the computer than to the stranger, even though that partner was not known. In our study, zero social closeness is important because it allows us to explore the _intrinsic_ motivation to keep the promises, if humans keep their word because it is morally correct, we would anticipate that probability of breaking the promise was very low in the three partners and, particularly, the inference regarding the probability of breaking the promise to the computer would include zero.

The probability of breaking the promise without social closeness, although low, is clearly greater than zero, which provides evidence that contradicts, at least in degree, the hypothesis of _intrinsic_ motivation. We cannot exclude the possibility that the subjects keep promises mainly morally motivated, because even in the case of the computer, the subjects keep a large proportion, but we can exclude that this is their only motivation.

On the other hand, high social closeness allows us to explore the _instrumental_ motivation to keep promises. In our study, it is almost 9 times more likely to break the promise to the stranger than to the friend; If human beings keep their word to facilitate cooperation in the future, we would anticipate that with a partner of high social closeness (with whom they are very likely to cooperate, even after the experiment) there would be less chance of breaking the promise. In fact, in our study, the promise to friends was not broken on any occasion, so the inference, in this case, is obvious.

Our findings could be explained by the social norm called _conditional cooperation_, which indicates that the belief that other people cooperate at high levels also induces high levels of cooperation [@Fehr2018]. Thus, the lack of certainty regarding the decisions that the computer would make could explain the probability of breaking the promise that exists towards this partner. Similarly, the information that the subjects have regarding the behavior of the friend -even before the experiment- can explain the high levels of cooperation and keeping of promises towards this partner.

Regarding cooperation, it is not the first investigation in which the effect of social closeness is evaluated, however, the nature of experimental manipulation has been diverse. For example, in one study the contribution in a public goods game was evaluated during several trials in a group made up of the same individuals (_partners_), compared to another group of new different subjects for each trial (_strangers_) [@Croson1996]. It was assumed that in the condition of _partners_ the cooperation would be greater because it is a stable group compared to the group of _strangers_. The mentioned study does not give details regarding the recruitment of participants, so we could assume that even in the condition of _partners_ these are individuals who do not consider themselves socially close.

In another experiment, an increase in cooperation in a trust game in pairs was reported when individuals are socially close [@Glaeser2000], in this study, subjects know each other and the researchers carefully measured several variables regarding their social connection, however, some individuals who arrived together were allowed to perform the task between themselves and others were paired with unclear criteria. The above allows social closeness between partners to be heterogeneous, allows for romantic relationships and assumes that two individuals who come together to class are considered close to each other. Also, since it is a study of _one-shot_, it excludes the possibility of evaluating how the same subject varies his behavior based on different levels of social closeness.

To try to homogenize social closeness between our subjects and their partners, participants performed the task with a same-sex match friend considered close by themselves, in none case was it a relative and, although we understand that there may be same-sex couples and that the subjects may have hidden their relationship, they were emphasized that to be part of the study their companion could not be a romantic or sexual partner. Besides, Bayesian analysis of the IOS scale showed evidence in favor that the degree of social closeness between 30 of our 45 subjects and their friends was practically the same. Although the results of the two studies mentioned are practically in the same direction as the findings of our experiment, establishing the similarity in effects magnitudes is not possible due to the design differences, the performed tasks, and the statistical procedures used.

A methodological contribution of our study is that, in the literature of promises, it is the first to use Bayesian inference tools: *Hierarchical Models* and *Cumulative Models*. The former (aka multilevel models), allowed us to model the effect of the experimental conditions on the individual response, these models assume that lower level observations (e.g. decisions of each individual) are nested in higher level units (e.g. individual subjects). Within-subjects designs have traditionally been analyzed with repeated measures AN(C)OVA, however, *Hierarchical Models* grant the advantages of naturally dealing with unbalanced data, including categorical and/or numerical predictors, explicitly incorporate individual variability, among others [@Vuorre2018; @Gelman2006; @McElreath2018; @Burkner2017].

In our case, as shown in Figure 5, the *Hierarchical Model* allowed us to capture very precisely how the experimental conditions affected the decisions of each subject and the differences between them. For example, there are obviously motivated _instrumental_ subjects such as number 8 or 12 who are very sensitive to the identity of their investors and modify their cooperation under social closeness. At the same time, there are notably _intrinsic_ subjects such as 9 and 10 who cooperate all the time regardless of who their partners are. The *Hierarchical Model* naturally includes this information for the estimation of population effects, which, if not considered, would lead to inaccurate inferences.

On the other hand, the *Cumulative Models* allowed us to model the strength of the commitment expressed through promises. A very frequent problem that has been pointed out recently is that analyzing ordinal data with methods that assume that observations are _metric_ can lead to serious inference errors [@Liddell2018]. The *Cumulative Models* assume that the observed responses come from the categorization of a continuous latent variable [@Burkner2019], so in our study, we could identify that the groups with different payment proportions, which in other articles have been called _dishonest_ and _honest_, differ in the commitment they express through their promises. Considering that the tools to perform the *Cumulative Models* are relatively recent, it would be valuable to replicate published studies [e.g., @Baumgartner2009; @Baumgartner2013] considering the new alternatives of inference.

Although the *Hierarchical* and *Cumulative* models are not tools exclusively for Bayesian inference, their application from this approach represents several advantages compared to frequentist statistics. Classic problems of the previous one such as multiple comparisons, the decision of when to stop collecting subjects (stopping rule), or use of planned comparisons versus _post hoc_, are not factors that affect the support provided by the data to the hypothesis from the Bayesian approach [@Dienes2011]. In Bayesian inference, hypothesis testing makes formal use of probability to express the plausibility of theories, in our case we were able to obtain evidence ratios regarding the extent to which our data support our hypotheses, regardless of whether these were proposed _a priori_ or _post hoc_.

A limitation of this paper is the use of hypothetical rewards compared to real rewards. A relatively recent study on decision-making reported that there is more loss aversion when subjects have real monetary rewards compared to hypothetical in a risk task [@Xu2016]. However, other works do not report differences between the use of both types of rewards in self-control, temporal and social discount tasks [@Locey2011; @Johnson2006; @Rachlin2006]. Likewise, it can be argued that the results have theoretical congruence [@Fehr2018] and are in the same direction as other works that use real money [@Croson1996; @Glaeser2000], so there are not many reasons to expect that other types of rewards would modify our results.

Another possible limitation is the use of multiple trials with each partner. Although the decision to use repeated measures during our design serves more the purpose of reducing contaminant sources and increasing internal validity [@Maxwell2018], to have more clarity regarding the difference in the _intrinsic_ and _instrumental_ motivations, the studies could benefit from only include one trial for each partner. Play a _one-shot_ with each partner avoids the possibility that multiple trials could generate the belief that the computer can also vary its behavior according to the decisions of the _trustee_.

# Conclusions

Subjects keep their promises by a combination between the _intrinsic_ and _instrumental_ motivations. They are predominantly _nice guys_ because they keep their word even with investors with no social closeness. However, its trustworthiness is far from perfect, since there is a small, but reliably greater than zero, proportion of betrayals committed towards this partner. It seems that if subjects cannot assure that their investor will be a _potential cooperator_, dishonesty may emerge in the form of broken promises. The theoretical predictions of the social norm _conditional cooperation_ frame the finding that social closeness decreases the probability of breaking the promise and increases cooperation. In the same way that suggests the correspondence between commitment expressed through promises and subsequent behavior.
